Template for indicator documentation

### 2023-09-19 DRAFT_NDVI_seminat.Rmd
ndvi wetland, seminat & natopen finished

@@

# Which ecological relevant variable(s) is the condition measure in the NiN data related to?
```{r warning=FALSE}
nin.seminat2 %>%
  filter(hovedtype=='T31') %>%
  filter(NiN_variable_code %in% c("7JB-BT","7RA-BA","7FA","7TK")) %>%
  mutate(NiN_variable_des=recode(NiN_variable_code,
                                 "7JB-BT"="Grazing pressure",
                                 "7RA-BA"="Succession",
                                 "7FA"="Alien species",
                                 "7TK"="Heavy vehicles")) %>%
  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
  geom_point(size=2) +  
  geom_jitter(width = 0.25, height=0.25) +
  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
  theme(legend.position="none") +
  ggtitle("Condition variables in boreal heath (T31)") +
  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")

nin.seminat2 %>%
  filter(hovedtype=='T32') %>%
  filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7SE","7JB-GJ")) %>%
  mutate(NiN_variable_des=recode(NiN_variable_code,
                                 "7JB-BA"="Land-use intensity",
                                 "7RA-SJ"="Succession",
                                 "7FA"="Alien species",
                                 "7SE"="Wear/tear",
                                 "7JB-GJ"="Fertilization")) %>%
  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
  geom_point(size=2) +  
  geom_jitter(width = 0.25, height=0.25) +
  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
  theme(legend.position="none") +
  ggtitle("Condition variables in semi-natural grasslands (T32)") +
  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")

nin.seminat2 %>%
  filter(hovedtype=='T33') %>%
  filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7JB-GJ")) %>%
  mutate(NiN_variable_des=recode(NiN_variable_code,
                                 "7JB-BA"="Land-use intensity",
                                 "7RA-SJ"="Succession",
                                 "7FA"="Alien species",
                                 "7JB-GJ"="Fertilization")) %>%
  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
  geom_point(size=2) +  
  geom_jitter(width = 0.25, height=0.25) +
  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
  theme(legend.position="none") +
  ggtitle("Condition variables in semi-natural coastal grasslands (T33)") +
  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")

nin.seminat2 %>%
  filter(hovedtype=='T34') %>%
  filter(NiN_variable_code %in% c("7JB-BT","7RA-SJ","7FA","7TK")) %>%
  mutate(NiN_variable_des=recode(NiN_variable_code,
                                 "7JB-BT"="Grazing pressure",
                                 "7RA-SJ"="Succession",
                                 "7FA"="Alien species",
                                 "7TK"="Heavy vehicles")) %>%
  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
  geom_point(size=2) +  
  geom_jitter(width = 0.25, height=0.25) +
  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
  theme(legend.position="none") +
  ggtitle("Condition variables in coastal heathlands (T34)") +
  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")

nin.seminat2 %>%
  filter(hovedtype %in% c('T40','T41')) %>%
  filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7SE","7JB-GJ")) %>%
  mutate(NiN_variable_des=recode(NiN_variable_code,
                                 "7JB-BA"="Land-use intensity",
                                 "7RA-SJ"="Succession",
                                 "7FA"="Alien species",
                                 "7SE"="Wear/tear",
                                 "7JB-GJ"="Fertilization")) %>%
  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
  geom_point(size=2) +  
  geom_jitter(width = 0.25, height=0.25) +
  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
  theme(legend.position="none") +
  ggtitle("Condition variables in previously heavily altered grasslands (T40/T41)") +
  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")

@@

We first investigate a model testing NDVI as a function of year and main ecosystem type for polygons in good condition, and we do so separately for each satellite.
```{r message=FALSE, warning=FALSE}
model.seminat.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$tilstand=="God",])
model.seminat.time.GodTilst.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$tilstand=="God",])
model.seminat.time.GodTilst.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$tilstand=="God",])

@@

Given the consistently higher NDVI-values with decreasing condition in boreal heath, We may ask the question if the temporal NDVI increase might be stronger under reduced condition?
```{r message=FALSE, warning=FALSE}
model.seminat.time.tilst.T31.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T31"),])
model.seminat.time.tilst.T31.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("T31"),])
model.seminat.time.tilst.T31.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("T31"),])
summary(model.seminat.time.tilst.T31.Sent)$coefficients$cond
summary(model.seminat.time.tilst.T31.Modi)$coefficients$cond
summary(model.seminat.time.tilst.T31.Land)$coefficients$cond
```
The temporal NDVI increase in boreal heath has NOT been stronger with reduced ecological condition. So a temporal NDVI approach does not seem like a promising way to go.
We check for the same pattern in T32 and T34, jsut because they are the other semi-natural major ecosystem types with a substantial number of occurrences in the the NiN data:
```{r message=FALSE, warning=FALSE}
model.seminat.time.tilst.T32.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T32"),])
model.seminat.time.tilst.T32.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("T32"),])
model.seminat.time.tilst.T32.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("T32"),])
summary(model.seminat.time.tilst.T32.Sent)$coefficients$cond
summary(model.seminat.time.tilst.T32.Modi)$coefficients$cond
summary(model.seminat.time.tilst.T32.Land)$coefficients$cond

@@

# 1) We are lacking ecosystem maps for semi-natural ecosystems
The study at hand clearly shows that different semi-natural ecosystem types display different greenness levels and thus need to be evaluated separately. Thus, we need ecosystem maps, that allow us to treat NDVI values, and temporal changes of these, for at least the different major ecosystem types within semi-natural systems separately.
# 2) We are lacking a good ecological understanding and ground-truthing for NDVI in semi-natural ecosystems
Unlike for ecosystems like forests, tundra, savannah and similar, there is little empirical work and ground-truthing data documented that link NDVI values to ecological phenomena in semi-natural systems. Here, we need field studies investigating more thoroughly how NDVI and ecological drivers, like for example shrub-/tree encroachment, are connected.

>>>>

# Where are the seminat sites?
```{r}
tm_shape(regnor) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_shape(SentinelNDVI.seminat) +
  tm_dots('hovedtype',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE), scale=2.5, legend.show = FALSE) + 
  tm_add_legend(type = "symbol", 
                col = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),
                labels = c("T31 - boreal heath",
                           "T32 - semi-natural grasslands",
                           "T33 - semi-natural coastal grasslands",
                           "T34 - coastal heathlands",
                           "T40/41 - previously heavily altered grassland"),
                title = "Major ecosystem type",
                size=3) +
  tm_layout(main.title = "Semi-natural ecosystems (NiN, Miljødirektoratets instruks)",
            main.title.size=1.2, legend.position = c("right", "bottom"), 
            legend.text.size = 1.3, legend.title.size = 1.4)
```

@@

# Which ecological relevant variable(s) is the condition measure in the NiN data related to?
```{r warning=FALSE}
nin.seminat2 %>%
  filter(hovedtype=='T31') %>%
  filter(NiN_variable_code %in% c("7JB-BT","7RA-BA","7FA","7TK")) %>%
  mutate(NiN_variable_des=recode(NiN_variable_code,
                                 "7JB-BT"="Grazing pressure",
                                 "7RA-BA"="Succession",
                                 "7FA"="Alien species",
                                 "7TK"="Heavy vehicles")) %>%
  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
  geom_point(size=2) +  
  geom_jitter(width = 0.25, height=0.25) +
  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
  theme(legend.position="none") +
  ggtitle("Condition variables in boreal heath (T31)") +
  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")

nin.seminat2 %>%
  filter(hovedtype=='T32') %>%
  filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7SE","7JB-GJ")) %>%
  mutate(NiN_variable_des=recode(NiN_variable_code,
                                 "7JB-BA"="Land-use intensity",
                                 "7RA-SJ"="Succession",
                                 "7FA"="Alien species",
                                 "7SE"="Wear/tear",
                                 "7JB-GJ"="Fertilization")) %>%
  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
  geom_point(size=2) +  
  geom_jitter(width = 0.25, height=0.25) +
  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
  theme(legend.position="none") +
  ggtitle("Condition variables in semi-natural grasslands (T32)") +
  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")

nin.seminat2 %>%
  filter(hovedtype=='T33') %>%
  filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7JB-GJ")) %>%
  mutate(NiN_variable_des=recode(NiN_variable_code,
                                 "7JB-BA"="Land-use intensity",
                                 "7RA-SJ"="Succession",
                                 "7FA"="Alien species",
                                 "7JB-GJ"="Fertilization")) %>%
  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
  geom_point(size=2) +  
  geom_jitter(width = 0.25, height=0.25) +
  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
  theme(legend.position="none") +
  ggtitle("Condition variables in semi-natural coastal grasslands (T33)") +
  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")

nin.seminat2 %>%
  filter(hovedtype=='T34') %>%
  filter(NiN_variable_code %in% c("7JB-BT","7RA-SJ","7FA","7TK")) %>%
  mutate(NiN_variable_des=recode(NiN_variable_code,
                                 "7JB-BT"="Grazing pressure",
                                 "7RA-SJ"="Succession",
                                 "7FA"="Alien species",
                                 "7TK"="Heavy vehicles")) %>%
  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
  geom_point(size=2) +  
  geom_jitter(width = 0.25, height=0.25) +
  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
  theme(legend.position="none") +
  ggtitle("Condition variables in coastal heathlands (T34)") +
  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")

nin.seminat2 %>%
  filter(hovedtype %in% c('T40','T41')) %>%
  filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7SE","7JB-GJ")) %>%
  mutate(NiN_variable_des=recode(NiN_variable_code,
                                 "7JB-BA"="Land-use intensity",
                                 "7RA-SJ"="Succession",
                                 "7FA"="Alien species",
                                 "7SE"="Wear/tear",
                                 "7JB-GJ"="Fertilization")) %>%
  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
  geom_point(size=2) +  
  geom_jitter(width = 0.25, height=0.25) +
  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
  theme(legend.position="none") +
  ggtitle("Condition variables in previously heavily altered grasslands (T40/T41)") +
  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")

@@

We first investigate a model testing NDVI as a function of year and main ecosystem type for polygons in good condition, and we do so separately for each satellite.
```{r, warning=FALSE}
model.seminat.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$tilstand=="God",])
model.seminat.time.GodTilst.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$tilstand=="God",])
model.seminat.time.GodTilst.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$tilstand=="God",])

@@

Given the consistently higher NDVI-values with decreasing condition in boreal heath, We may ask the question if the temporal NDVI increase might be stronger under reduced condition?
```{r, warning=FALSE}
model.seminat.time.tilst.T31.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T31"),])
model.seminat.time.tilst.T31.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("T31"),])
model.seminat.time.tilst.T31.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("T31"),])
summary(model.seminat.time.tilst.T31.Sent)$coefficients$cond
summary(model.seminat.time.tilst.T31.Modi)$coefficients$cond
summary(model.seminat.time.tilst.T31.Land)$coefficients$cond
```
The temporal NDVI increase in boreal heath has NOT been stronger with reduced ecological condition. So a temporal NDVI approach does not seem like a promising way to go.
We check for the same pattern in T32 and T34, jsut because they are the other semi-natural major ecosystem types with a substantial number of occurrences in the the NiN data:
```{r, warning=FALSE}
model.seminat.time.tilst.T32.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T32"),])
model.seminat.time.tilst.T32.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("T32"),])
model.seminat.time.tilst.T32.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("T32"),])
summary(model.seminat.time.tilst.T32.Sent)$coefficients$cond
summary(model.seminat.time.tilst.T32.Modi)$coefficients$cond
summary(model.seminat.time.tilst.T32.Land)$coefficients$cond

@@

# 1) We are lacking ecosystem maps for semi-natural ecosystems
The analysis at hand clearly shows that different semi-natural ecosystem types display different greenness levels and thus need to be evaluated separately. Thus, we need ecosystem maps, that allow us to treat NDVI values, and temporal changes of these, for at least the different major ecosystem types within semi-natural systems separately.
# 2) We are lacking a good ecological understanding and ground-truthing for NDVI in semi-natural ecosystems
Unlike for ecosystems like forests, tundra, savannah and similar, there is little empirical work and ground-truthing data documented that link NDVI values to ecological phenomena in semi-natural systems. Here, we need field studies investigating more thoroughly how NDVI and ecological drivers, like for example shrub-/tree encroachment, are connected. This includes a need for understanding temporal increases in NDVI in ecosystems in good condition.

### 2023-09-19 DRAFT_NDVI_wetland.Rmd
ndvi wetland, seminat & natopen finished

@@

# Which ecological relevant variable(s) is the condition measure in the NiN data related to?
```{r}
nin.wetland2 %>%
  filter(hovedtype=='V9') %>%
  filter(NiN_variable_code %in% c("1AG-B","7GR-GI","1AG-A-G","1AR-C-L","7TK","PRSL")) %>%
  mutate(NiN_variable_des=recode(NiN_variable_code,
                                 "1AG-B"="Shrub cover",
                                 "7GR-GI"="Abundance of ditches",
                                 "1AG-A-G"="Tree cover",
                                 "1AR-C-L"="Woody plants in field layer",
                                 "7TK"="Heavy vehicles",
                                 "PRSL"="Wear/tear")) %>%
  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
  geom_point(size=2) +  
  geom_jitter(width = 0.25, height=0.25) +
  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
  theme(legend.position="none") +
  ggtitle("Condition variables in semi-natural mires (V9)") +
  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")

nin.wetland2 %>%
  filter(hovedtype=='V10') %>%
  filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7JB-GJ")) %>%
  mutate(NiN_variable_des=recode(NiN_variable_code,
                                 "7JB-BA"="Land use intensity",
                                 "7RA-SJ"="Succession",
                                 "7FA"="Alien species",
                                 "7JB-GJ"="Fertilization")) %>%
  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
  geom_point(size=2) +  
  geom_jitter(width = 0.25, height=0.25) +
  facet_wrap(~NiN_variable_des,nrow=2, ncol=2) +
  theme(legend.position="none") +
  ggtitle("Condition variables in semi-natural wet-meadows (V10)") +
  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")

nin.wetland2 %>%
  filter(hovedtype=='V1') %>%
  filter(NiN_variable_code %in% c("7GR-GI","PRTK","7FA","7TK","7SE","PRSL")) %>%
  mutate(NiN_variable_des=recode(NiN_variable_code,
                                 "7GR-GI"="Abundance of ditches",
                                 "PRTK"="Heavy vehicles",
                                 "7FA"="Alien species",
                                 "7TK"="Heavy vehicles2",
                                 "7SE"="Wear/tear",
                                 "PRSL"="Wear/tear2")) %>%
  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
  geom_point(size=2) +  
  geom_jitter(width = 0.25, height=0.25) +
  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
  theme(legend.position="none") +
  ggtitle("Condition variables in minerotrophic fens (V1)") +
  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")

@@

First, we take a look at how balanced the data are across major ecosystem types, regions and condition classes
```{r warning=FALSE, message=FALSE}
# NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
SentinelNDVI.wetland %>%

@@

We first investigate a model testing NDVI as a function of year and main ecosystem type for polygons in good condition, and we do so separately for each satellite.
```{r message=FALSE, warning=FALSE}
model.wetland.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.wetland[SentinelNDVI.wetland$tilstand=="God",])
model.wetland.time.GodTilst.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.wetland[ModisNDVI.wetland$tilstand=="God",])
model.wetland.time.GodTilst.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.wetland[LandsatNDVI.wetland$tilstand=="God",])

@@

Given the systematically higher NDVI-values with decreasing condition in semi-natural wetlands, We may ask the question if the temporal NDVI increase might be stronger with reduced condition?
```{r message=FALSE, warning=FALSE}
model.wetland.time.tilst.V9V10.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.wetland[SentinelNDVI.wetland$hovedtype %in% c("V9","V10"),])
model.wetland.time.tilst.V9V10.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.wetland[ModisNDVI.wetland$hovedtype %in% c("V9","V10"),])
model.wetland.time.tilst.V9V10.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.wetland[LandsatNDVI.wetland$hovedtype %in% c("V9","V10"),])
summary(model.wetland.time.tilst.V9V10.Sent)$coefficients$cond
summary(model.wetland.time.tilst.V9V10.Modi)$coefficients$cond
summary(model.wetland.time.tilst.V9V10.Land)$coefficients$cond
```
Indeed, the temporal NDVI increase has been stronger in semi-natural wetlands with reduced ecological condition, in both Sentinel and Landsat. For MODIS, the time slope is again 1/10 of the slopes in the other two satellite time series, and it is not significant.

@@

# 1) We are lacking ecosystem maps for wetland ecosystems
The study at hand clearly shows that different wetland ecosystem types display different greenness levels and thus need to be evaluated separately. Thus, we need ecosystem maps, that allow us to treat NDVI values, and temporal changes of these, for at least the different major ecosystem types within wetlands separately.
# 2) We are lacking a good ecological understanding and ground-truthing for NDVI in wetland ecosystems
Unlike for ecosystems like forests, tundra, savannah and similar, there is little empirical work and ground-truthing data documented that link NDVI values to ecological phenomena in wetlands. Here, we need field studies investigating:
- The connection between NDVI and ecological drivers like for example potential effects of Nitrogen deposition

>>>>

# Where are the wetland sites?
```{r}
tm_shape(regnor) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_shape(SentinelNDVI.wetland) +
  tm_dots('hovedtype',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 6, plot = FALSE), scale=2.5, legend.show = FALSE) + 
  tm_add_legend(type = "symbol", 
                col = tmaptools::get_brewer_pal("YlOrRd", 6, plot = FALSE),
                labels = c("V1 - minerotrophic mires",
                           "V3 - ombrotrophic bogs",
                           "V4 - cold springs",
                           "V6 - wet snowbeds",
                           "V9 - semi-natural mires",
                           "V10 - semi-natural wet meadows"),
                title = "Major ecosystem type",
                size=3) +
  tm_layout(main.title = "Wetland ecosystems (NiN, Miljødirektoratets instruks)",
            main.title.size=1.2, legend.position = c("right", "bottom"), 
            legend.text.size = 1.3, legend.title.size = 1.4)
```

@@

# Which ecological relevant variable(s) is the condition measure in the NiN data related to?
```{r}
nin.wetland2 %>%
  filter(hovedtype=='V9') %>%
  filter(NiN_variable_code %in% c("1AG-B","7GR-GI","1AG-A-G","1AR-C-L","7TK","PRSL")) %>%
  mutate(NiN_variable_des=recode(NiN_variable_code,
                                 "1AG-B"="Shrub cover",
                                 "7GR-GI"="Abundance of ditches",
                                 "1AG-A-G"="Tree cover",
                                 "1AR-C-L"="Woody plants in field layer",
                                 "7TK"="Heavy vehicles",
                                 "PRSL"="Wear/tear")) %>%
  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
  geom_point(size=2) +  
  geom_jitter(width = 0.25, height=0.25) +
  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
  theme(legend.position="none") +
  ggtitle("Condition variables in semi-natural mires (V9)") +
  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")

nin.wetland2 %>%
  filter(hovedtype=='V10') %>%
  filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7JB-GJ")) %>%
  mutate(NiN_variable_des=recode(NiN_variable_code,
                                 "7JB-BA"="Land use intensity",
                                 "7RA-SJ"="Succession",
                                 "7FA"="Alien species",
                                 "7JB-GJ"="Fertilization")) %>%
  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
  geom_point(size=2) +  
  geom_jitter(width = 0.25, height=0.25) +
  facet_wrap(~NiN_variable_des,nrow=2, ncol=2) +
  theme(legend.position="none") +
  ggtitle("Condition variables in semi-natural wet-meadows (V10)") +
  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")

nin.wetland2 %>%
  filter(hovedtype=='V1') %>%
  filter(NiN_variable_code %in% c("7GR-GI","PRTK","7FA","7TK","7SE","PRSL")) %>%
  mutate(NiN_variable_des=recode(NiN_variable_code,
                                 "7GR-GI"="Abundance of ditches",
                                 "PRTK"="Heavy vehicles",
                                 "7FA"="Alien species",
                                 "7TK"="Heavy vehicles2",
                                 "7SE"="Wear/tear",
                                 "PRSL"="Wear/tear2")) %>%
  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
  geom_point(size=2) +  
  geom_jitter(width = 0.25, height=0.25) +
  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
  theme(legend.position="none") +
  ggtitle("Condition variables in minerotrophic fens (V1)") +
  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")

@@

First, we take a look at how balanced the data are across major ecosystem types, regions and condition classes
```{r warning=FALSE}
# NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
SentinelNDVI.wetland %>%

@@

We first investigate a model testing NDVI as a function of year and main ecosystem type for polygons in good condition, and we do so separately for each satellite.
```{r warning=FALSE}
model.wetland.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.wetland[SentinelNDVI.wetland$tilstand=="God",])
model.wetland.time.GodTilst.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.wetland[ModisNDVI.wetland$tilstand=="God",])
model.wetland.time.GodTilst.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.wetland[LandsatNDVI.wetland$tilstand=="God",])

@@

Given the systematically higher NDVI-values with decreasing condition in semi-natural wetlands, We may ask the question if the temporal NDVI increase might be stronger with reduced condition?
```{r warning=FALSE}
model.wetland.time.tilst.V9V10.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.wetland[SentinelNDVI.wetland$hovedtype %in% c("V9","V10"),])
model.wetland.time.tilst.V9V10.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.wetland[ModisNDVI.wetland$hovedtype %in% c("V9","V10"),])
model.wetland.time.tilst.V9V10.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.wetland[LandsatNDVI.wetland$hovedtype %in% c("V9","V10"),])
summary(model.wetland.time.tilst.V9V10.Sent)$coefficients$cond
summary(model.wetland.time.tilst.V9V10.Modi)$coefficients$cond
summary(model.wetland.time.tilst.V9V10.Land)$coefficients$cond
```
Indeed, the temporal NDVI increase has been stronger in semi-natural wetlands with reduced ecological condition, in both Sentinel and Landsat. For MODIS, the time slope is again 1/10 of the slopes in the other two satellite time series, and it is not significant.

@@

# 1) We are lacking ecosystem maps for wetland ecosystems
The analysis at hand clearly shows that different wetland ecosystem types display different greenness levels and thus need to be evaluated separately. Thus, we need ecosystem maps, that allow us to treat NDVI values, and temporal changes of these, for at least the different major ecosystem types within wetlands separately.
# 2) We are lacking a good ecological understanding and ground-truthing for NDVI in wetland ecosystems
Unlike for ecosystems like forests, tundra, savannah and similar, there is little empirical work and ground-truthing data documented that link NDVI values to ecological phenomena in wetlands. Here, we need field studies investigating:
- The connection between NDVI and ecological drivers like for example potential effects of Nitrogen deposition. This includes a need for understanding temporal increases in NDVI in ecosystems in good condition.
